---
title: "CM 2605: Simulation and Modeling Techniques"
subtitle: "CW Part 01 — Simulation and Stochastic Analysis of Hospital Emergency Resources"
author: "J. A. N. Vishara Jayalath | RGU: 2425578 | IIT: 20240184"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Simulation and Stochastic Analysis of Hospital Emergency Resources

# Introduction

This report presents a discrete-event simulation of the Emergency Room (ER) triage process at a hospital experiencing overcrowding during its weekend shifts. The simulation models patient arrivals and service provided using appropriate stochastic distributions, and evaluates whether the current staffing level of 3 triage doctors meets the hospital board's mandated threshold of an average patient waiting time not exceeding 10 minutes.

A second scenario is also simulated with 4 triage doctors to assess and compare whether the additional resource resolves the overcrowding issue.

---

# Model Parameters

The system is modelled as an **M/M/c queue**, where:

- Arrivals follow a **Poisson process** with rate **$\lambda = 12$** patients/hour ($= 0.2$ patients/minute)
- Service times follow an **Exponential distribution** with mean $\mu = 12$ minutes per patient per doctor
- Queue discipline is **First-Come, First-Served (FCFS)**
- Number of servers $c = 3$ (Scenario 1) and $c = 4$ (Scenario 2)
- We simulate $n = 500$ patients

The server utilisation factor is:

$$\rho = \frac{\lambda}{c \cdot \mu^{-1}} = \frac{12 \text{ patients/hr}}{c \times 5 \text{ patients/hr/doctor}}$$
For $c = 3$: $\rho = \frac{12}{15} = 0.80$

For $c = 4$: $\rho = \frac{12}{20} = 0.60$

---

# Simulation Algorithm

## Helper Function: Multi-Server Queue Simulation

```{r}

# Multi- Server Queue Simulation (M/M/c)

# Arguments:
#   n_patients: number of patients to simulate
#   arrival_rate: average patients per minute (lambda)
#   service_rate: average service time in minutes per patient (1/mu)
#   seed: random seed for reproducibility of results

simulate_er <- function(n_patients = 500, 
                        arrival_time = 12/60, # patients per minute
                        service_mean = 12, # average minutes per patient
                        n_doctors = 3,
                        seed = 42){
  set.seed(seed)
  
  
  
  # Step 1: Generate inter-arrival times (Exponential from Poisson process)
  # If arrivals are Poisson (lambda), inter-arrival gaps are Exp(lambda)
  inter_arrivals <- rexp(n = n_patients, rate = arrival_rate)
  
  # Step 2: Compute cumulative arrival times
  arrival_times <- cumsum(inter_arrivals)
  
  # Step 3: Generate service durations for each patient 
  service_durations <- rexp(n = n_patients, rate = 1 / service_mean)
  
  # Step 4: Track when each doctor becomes free
  # Initialize all doctors as free from time 0
  doctor_free_at <- rep(0,n_doctors)
  
  # Step 5: Pre-allocate result vectors
  service_start <- numeric(n_patients)
  service_end <- numeric (n_patients)
  wait_time <- numeric (n_patients)
  system_time <- numeric (n_patients)
  
  # Step 6
}
```

















